Self-similarity of the elements of the feasible set

### 2020-08-13 analysis/reports/self_similarity.Rmd
draft self-similarity report

Here we have a bank of `r length(unique(example_fs$sim))` samples from the feasible set for SADs with 7 species and 71 individuals.

@@

ggplot(example_fs, aes(x = rank, y = abund, group = sim)) +
  geom_line(alpha = .1) +
  theme_bw()
```

We draw two of these samples to compare.

```{r draw two}

@@

```{r rep sampler, fig.dim = c(3,3)}

@@

```{r comparison 2}

@@

>>>>

Here we have a bank of `r length(unique(example_fs$sim))` samples from the feasible set for SADs with 7 species and 71 individuals. We draw two of these samples to compare.

@@

ggplot(example_fs, aes(x = rank, y = abund, group = sim)) +
  geom_line(alpha = .1) +
  theme_bw()

@@

```{r rep sampler, fig.dim = c(2,2)}

@@

```{r comparison 2, fig.dim = c(2,2)}

@@

### Across a range of S and N

Here we have drawn samples from a "net" of points in S and N space that spans the range present in our datasets. For each feasible set we make 200 comparisons of elements (although for small feasible sets, 200 is not necessarily possible). Here is how that net is distributed in S and N space, colored by the log() number of elements in the feasible set.

```{r diffs}

loadd(all_diffs, cache= cache)

all_diffs <- all_diffs %>%
  mutate(log_nparts = log(gmp:::as.double.bigz(nparts))) %>%
  filter(!is.na(sim1), !is.na(sim2))

ggplot(all_diffs, aes(log(s0), log(n0), color = log_nparts)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_c() +
  theme(legend.position = "top")

#### Heat maps

We can make heat maps of the density distribution of self-similarity metrics and look at how the densities shift over gradients in the size of the feasible set.

There are a couple of nuances to doing this:

* We need to have the same number of comparisons for every feasible set, so that the density distributions will be comparable. In order to get a large swatch of S by N space, I've taken 50 comparisons from every feasible set. This is pretty low, so it lets us get even small feasible sets. Setting the minimum higher does not change the overall impression.
* The metrics vary in how they are bounded. For some of them (R2, log r2) they can have nonsensical long tails towards very low values. For visualization, I've filtered out the long tail values (prior to selecting the 50 comparisons). The long tails tend to be most dominant in the comparisons made from small feasible sets. Removing them therefore makes the small feasible sets look *more* self-similar.

##### R2

Higher R2 values indicate more similarity. R2 can be *very low* but the most meaningful variation is between 0 and 1.

```{r heat maps}

loadd(all_diffs, cache= cache)

# For plotting remove r2 that are EXTREMELY LOW.

all_diffs_r2 <- all_diffs %>%
    mutate(log_nparts = log(gmp:::as.double.bigz(nparts))) %>%
  mutate(s_and_n = paste0("s_", s0, "_n_", n0),
         low_r2 = r2 <= -1) %>%
  group_by(s_and_n) %>%
  mutate(prop_small_r2 = mean(low_r2)) %>%
  filter(r2 > -1) %>%
  group_by(s_and_n) %>%
  mutate(ncomparisons = dplyr::n(),
         comparison_index = dplyr::row_number()) %>%
  ungroup() %>%
  filter(ncomparisons > 50, comparison_index <= 50) %>%
  group_by(s0, n0) %>%
  mutate(comparisons_included = dplyr::n()) %>%
  ungroup()

ggplot(all_diffs_r2, aes(log(s0), log(n0), color = log_nparts)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma") +
  theme(legend.position = "top")

# This plot demonstrates that removing comparisons for which r2 < -1 most impacts small FS. So if it distorts the results at all, it makes the small ones look *higher* than they really are.
# ggplot(all_diffs_r2, aes(log(s0), log(n0), color = prop_small_r2)) +
#   geom_point() +
#   theme_bw() +
#   scale_color_viridis_c() +
#   theme(legend.position = "top")

ggplot(all_diffs_r2, aes(r2, group = s_and_n, color = log_nparts)) +
  geom_density(alpha = .4) +
  theme_bw() +
  theme(legend.position = "top") +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma")

Here is R2 for only "small" feasible sets (those with fewer than `r exp(20)` elements in the feasible set). This lets us zoom in on the distributions where they start to broaden out.

```{r r2 small}

ggplot(filter(all_diffs_r2, log_nparts < 20), aes(log(s0), log(n0), color = log_nparts)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma") +
  theme(legend.position = "top")

ggplot(filter(all_diffs_r2, log_nparts < 20), aes(r2, group = s_and_n, color = log_nparts)) +
  geom_density(alpha = .4) +
  theme_bw() +
  theme(legend.position = "top") +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma")
```

##### R2 on logged vectors

Higher R2 values indicate more similarity. R2 can be *very low* but the most meaningful variation is between 0 and 1.

```{r r2 on log}

loadd(all_diffs, cache= cache)

# For plotting remove r2 that are EXTREMELY LOW.

all_diffs_r2_log <- all_diffs %>%
    mutate(log_nparts = log(gmp:::as.double.bigz(nparts))) %>%
  mutate(s_and_n = paste0("s_", s0, "_n_", n0),
         low_r2_log = r2_log <= -1) %>%
  group_by(s_and_n) %>%
  mutate(prop_small_r2_log = mean(low_r2_log)) %>%
  filter(r2_log > -1) %>%
  group_by(s_and_n) %>%
  mutate(ncomparisons = dplyr::n(),
         comparison_index = dplyr::row_number()) %>%
  ungroup() %>%
  filter(ncomparisons > 50, comparison_index <= 50) %>%
  group_by(s0, n0) %>%
  mutate(comparisons_included = dplyr::n()) %>%
  ungroup()

ggplot(all_diffs_r2_log, aes(log(s0), log(n0), color = log_nparts)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma") +
  theme(legend.position = "top")

# This plot demonstrates that removing comparisons for which r2 < -1 most impacts small FS. So if it distorts the results at all, it makes the small ones look *higher* than they really are.
# ggplot(all_diffs_r2, aes(log(s0), log(n0), color = prop_small_r2)) +
#   geom_point() +
#   theme_bw() +
#   scale_color_viridis_c() +
#   theme(legend.position = "top")

ggplot(all_diffs_r2_log, aes(r2_log, group = s_and_n, color = log_nparts)) +
  geom_density(alpha = .4) +
  theme_bw() +
  theme(legend.position = "top") +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma")

Here is R2 on logged vectors for only "small" feasible sets (those with fewer than `r exp(20)` elements in the feasible set). This lets us zoom in on the distributions where they start to broaden out.

```{r r2 log small}

ggplot(filter(all_diffs_r2_log, log_nparts < 20), aes(log(s0), log(n0), color = log_nparts)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma") +
  theme(legend.position = "top")

ggplot(filter(all_diffs_r2_log, log_nparts < 20), aes(r2_log, group = s_and_n, color = log_nparts)) +
  geom_density(alpha = .4) +
  theme_bw() +
  theme(legend.position = "top") +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma")
```

##### Coefficient of determination

Higher CD values indicate more similarity. It is bounded 0 to 1.

```{r }

loadd(all_diffs, cache= cache)

# For plotting remove r2 that are EXTREMELY LOW.

all_diffs_cd <- all_diffs %>%
    mutate(log_nparts = log(gmp:::as.double.bigz(nparts))) %>%
  mutate(s_and_n = paste0("s_", s0, "_n_", n0),
         low_cd = cd <= -1) %>%
  group_by(s_and_n) %>%
  mutate(prop_small_cd = mean(low_cd)) %>%
#  filter(cd > -1) %>%
  group_by(s_and_n) %>%
  mutate(ncomparisons = dplyr::n(),
         comparison_index = dplyr::row_number()) %>%
  ungroup() %>%
  filter(ncomparisons > 50, comparison_index <= 50) %>%
  group_by(s0, n0) %>%
  mutate(comparisons_included = dplyr::n()) %>%
  ungroup()

ggplot(all_diffs_cd, aes(log(s0), log(n0), color = log_nparts)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma") +
  theme(legend.position = "top")

# This plot demonstrates that removing comparisons for which cd < -1 most impacts small FS. So if it distorts the results at all, it makes the small ones look *higher* than they really are.
# ggplot(all_diffs_cd, aes(log(s0), log(n0), color = prop_small_cd)) +
#   geom_point() +
#   theme_bw() +
#   scale_color_viridis_c() +
#   theme(legend.position = "top")

ggplot(all_diffs_cd, aes(cd, group = s_and_n, color = log_nparts)) +
  geom_density(alpha = .4) +
  theme_bw() +
  theme(legend.position = "top") +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma")

Here is cd for only "small" feasible sets (those with fewer than `r exp(20)` elements in the feasible set). This lets us zoom in on the distributions where they start to broaden out.

```{r cd small}

ggplot(filter(all_diffs_cd, log_nparts < 20), aes(log(s0), log(n0), color = log_nparts)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma") +
  theme(legend.position = "top")

ggplot(filter(all_diffs_cd, log_nparts < 20), aes(cd, group = s_and_n, color = log_nparts)) +
  geom_density(alpha = .4) +
  theme_bw() +
  theme(legend.position = "top") +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma")
```

##### Proportion of individuals allocated to different species

Lower "prop off" values indicate more similarity. It is bounded 0 to 1.

```{r prop off heat maps}

loadd(all_diffs, cache= cache)

# For plotting remove r2 that are EXTREMELY LOW.

all_diffs_prop_off <- all_diffs %>%
    mutate(log_nparts = log(gmp:::as.double.bigz(nparts))) %>%
  mutate(s_and_n = paste0("s_", s0, "_n_", n0),
         low_prop_off = prop_off <= -1) %>%
  group_by(s_and_n) %>%
  mutate(prop_small_prop_off = mean(low_prop_off)) %>%
#  filter(prop_off > -1) %>%
  group_by(s_and_n) %>%
  mutate(ncomparisons = dplyr::n(),
         comparison_index = dplyr::row_number()) %>%
  ungroup() %>%
  filter(ncomparisons > 50, comparison_index <= 50) %>%
  group_by(s0, n0) %>%
  mutate(comparisons_included = dplyr::n()) %>%
  ungroup()

ggplot(all_diffs_prop_off, aes(log(s0), log(n0), color = log_nparts)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma") +
  theme(legend.position = "top")

# This plot demonstrates that removing comparisons for which prop_off < -1 most impacts small FS. So if it distorts the results at all, it makes the small ones look *higher* than they really are.
# ggplot(all_diffs_prop_off, aes(log(s0), log(n0), color = prop_small_prop_off)) +
#   geom_point() +
#   theme_bw() +
#   scale_color_viridis_c() +
#   theme(legend.position = "top")

ggplot(all_diffs_prop_off, aes(prop_off, group = s_and_n, color = log_nparts)) +
  geom_density(alpha = .4) +
  theme_bw() +
  theme(legend.position = "top") +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma")

Here is prop_off for only "small" feasible sets (those with fewer than `r exp(20)` elements in the feasible set). This lets us zoom in on the distributions where they start to broaden out.

```{r prop_off small}

ggplot(filter(all_diffs_prop_off, log_nparts < 20), aes(log(s0), log(n0), color = log_nparts)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma") +
  theme(legend.position = "top")

ggplot(filter(all_diffs_prop_off, log_nparts < 20), aes(prop_off, group = s_and_n, color = log_nparts)) +
  geom_density(alpha = .4) +
  theme_bw() +
  theme(legend.position = "top") +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma")
```

##### K-L divergence

Lower divergence  values indicate more similarity. It is bounded 0 to 1.

```{r div heat maps}

loadd(all_diffs, cache= cache)

# For plotting remove r2 that are EXTREMELY LOW.

all_diffs_div <- all_diffs %>%
    mutate(log_nparts = log(gmp:::as.double.bigz(nparts))) %>%
  mutate(s_and_n = paste0("s_", s0, "_n_", n0),
         low_div = div <= -1) %>%
  group_by(s_and_n) %>%
  mutate(prop_small_div = mean(low_div)) %>%
#  filter(div > -1) %>%
  group_by(s_and_n) %>%
  mutate(ncomparisons = dplyr::n(),
         comparison_index = dplyr::row_number()) %>%
  ungroup() %>%
  filter(ncomparisons > 50, comparison_index <= 50) %>%
  group_by(s0, n0) %>%
  mutate(comparisons_included = dplyr::n()) %>%
  ungroup()

ggplot(all_diffs_div, aes(log(s0), log(n0), color = log_nparts)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma") +
  theme(legend.position = "top")

# This plot demonstrates that removing comparisons for which div < -1 most impacts small FS. So if it distorts the results at all, it makes the small ones look *higher* than they really are.
# ggplot(all_diffs_div, aes(log(s0), log(n0), color = prop_small_div)) +
#   geom_point() +
#   theme_bw() +
#   scale_color_viridis_c() +
#   theme(legend.position = "top")

ggplot(all_diffs_div, aes(div, group = s_and_n, color = log_nparts)) +
  geom_density(alpha = .4) +
  theme_bw() +
  theme(legend.position = "top") +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma")

Here is div for only "small" feasible sets (those with fewer than `r exp(20)` elements in the feasible set). This lets us zoom in on the distributions where they start to broaden out.

```{r div small}

ggplot(filter(all_diffs_div, log_nparts < 20), aes(log(s0), log(n0), color = log_nparts)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma") +
  theme(legend.position = "top")

ggplot(filter(all_diffs_div, log_nparts < 20), aes(div, group = s_and_n, color = log_nparts)) +
  geom_density(alpha = .4) +
  theme_bw() +
  theme(legend.position = "top") +
  scale_color_viridis_c(direction = -1, end = .8, option = "plasma")
```
